Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden

### 2023-09-10 analysis/paper/paper.qmd
Added table. Updated to pdf output, referencing tables requires this.

---
title: "Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
  - Johan Linderholm:
      email: johan.linderholm@umu.se
      institute: [UmU]
      correspondence: false
  - Paul Geladi:
      email: paul.geladi@slu.se
      institute: [SLU]
      correspondence: false
  - Philip I. Buckland:
      email: philip.buckland@umu.se
      institute: [UmU]
      correspondence: false
institute:
  - UmU: Environmental Archaeology Lab, Department of Historical, Philosophical and Religious studies, Umeå University, Umeå, Sweden
  - SLU: Department of Forest Biomaterials and Technology, Swedish University of Agricultural Sciences, Umeå, Sweden 
title-block-published: "Last updated"  
date: now
date-format: long
crossref: 
  fig-title: '**Fig.**'
  fig-labels: '**arabic**'
  fig-prefix: Fig.
format:
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 300
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/quaternary-science-reviews.csl" # Insert path for the bib-style
abstract: |
  Northern Fennoscandia represents a geologically complex region affected by both glacial and postglacial processes. Provenance of locally available geological material thus becomes complex and fraught with uncertainty. Quartzite was a key material type utilized by hunter-gatherers in Northern Sweden around the 4 000 – 2 000 BP period, and is thus critical to the understanding of raw material procurement and material flow within the region. However, there is a severe lack of methodological development in the characterization of these materials. 125 points and preforms made from quartz/quartzite were sampled from 46 archaeological sites along the upper Ångerman river valley in Västerbotten, Sweden. The material has been analysed non-destructively using three separate portable spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence). Evaluation of the spectra and exploratory data analysis using Principal Component Analysis demonstrates detectable differences in the material that likely stem from diagenetic/paragenetic origin. The presence of graphite, muscovite and biotite could likewise provide information on the material’s metamorphic grade. In addition to reaffirming the potential of field-based screening instrumentation, these results will benefit future surveys of geological sources in the region. They also indicate potential for the construction of a predictive model that could classify the quartzite based on its chemical characteristic. Such a model would prove useful in future spatial analysis and testing of models of raw material management.
keywords: |
  Hunter-Gatherer; Lithics; Provenance; Quartzite; Vibrational spectroscopy; Scandinavia
---

@@

<!-- The following code chunk defines some general settings how code chunks should behave. -->
```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300,
  encoding = "UTF-8"
)
```

@@

This study includes spectroscopic surface measurements collected on bifacial points (probable arrowheads and spear points) made of quartz and quartzite. The points have been recovered from archaeological sites during excavations and surveys, including stray finds, in Västerbotten County, Sweden (@fig-distribution-map). The sites are distributed along the Ångerman river valley, from Lake Kultsjön in the north, towards the town of Åsele in the south. The dataset consists of `r length(tot_uniq_samples)` points, point fragments and preforms stored in the collections at Västerbottens museum (www.vbm.se) and have been analysed using NIRS, Raman scattering and XRF.

@@

```{r}
#| label: fig-distribution-map
#| cache: false
#| cache-lazy: false
#| fig-cap: "Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden. The glacial disturbance is visible in the widely distributed moraine. Tectonic symbology after [@gee_stephens20]. Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet"

knitr::include_graphics(here::here("analysis", "figures", "site_distribution_v1.jpg"), error = FALSE)
```

@@

@@

```{r}
#| label: fig-artefacts
#| cache: false
#| cache-lazy: false
#| fig-cap: "Examples of the bifacial points studied. The material is heterogeneous in terms of colour, and the surface of the artefacts themselves is not uniform. Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation. Length varies between 12 – 138 mm, width between 8 – 67 mm, and thickness between 3 – 76 mm"

knitr::include_graphics(here::here("analysis", "figures", "figure_2.jpg"), error = FALSE)
```

@@

```{r}
#| label: fig-nir-peak-summary
#| cache: true
#| fig-width: 7.87
#| fig-asp: 1
#| out-width: "70%"
#| fig-align: center
#| fig-cap: "The main features in the NIR spectra can be related to overtones and vibration combinations of OH groups. The 1 400 nm and 1 900 nm peaks are indicative of molecular water, and common in quartz-based materials. The 2 200 nm and 2 350 nm bands indicate metal-OH combinations, where a more intense 2 200 nm band has typically been related to the presence of AlOH, and a more intense 2 350 nm band MgOH. Although difficult to detect in the non-processed spectra, weaker features also appear at 1 065 nm and 1 300 nm, that may be the result of ferrous iron (Fe^2+^)"

@@

```{r}
#| label: fig-nir-pca
#| cache: true
#| fig-width: 7.87
#| fig-asp: 1
#| out-width: "70%"
#| fig-align: center
#| fig-cap: "NIR PCA score plot for PC 1-3: a) dark and light material is separated out along PC 1, due to absorbance, with white and colourless material separating along PC 2, b) PC 3 adds little new information, mainly separating three of the light samples"
#| 
quietly(source(here::here("analysis", "paper","004-nir-pca.R"), encoding = "UTF-8"))
#knitr::include_graphics(here::here("analysis", "figures", "004-nir-pca-with-photos.jpg"), error = FALSE)

@@

\linebreak
<!--
```{r}
#| label: fig-raman-spectra-summary
#| cache: true
#| fig-width: 7.87
#| fig-asp: 0.8
#| out-width: "70%"
#| fig-align: center
#| fig-cap: "The non-processed Raman spectra visualised by colour group. Note that the dark material in general features more intense peaks at wavenumber ~ 1 300 - 1 600 (cm^-1^)"

@@

```{r}
#| label: fig-raman-peak-ident
#| cache: true
#| fig-width: 7.87
#| fig-asp: 0.8
#| out-width: "70%"
#| fig-align: center
#| fig-cap: "TAveraged and baseline corrected Raman spectra for the light and dark groups. The features can be related to quartz and carbon structures in the material. The initial three peaks in 100 - 500 cm^-1^ range correspond to some of the more well-known vibration modes of quartz. The 1 287 cm^-1^ and 1 598 cm^-1^ peaks fall within what is referred to as the D and G band respectively, indicating the presence of carbon structures"

@@

```{r}
#| label: fig-raman-pca
#| cache: false
#| cache-lazy: false
#| fig-cap: "PCA score plot for Raman data. Objects that feature only quartz peaks are clustered towards the top left of the plot, whilst the darker material that feature dominant D/G peaks are distributed towards the right of the plot. In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster"

#quietly(source(here::here("analysis", "paper", "007-raman-pca.R"), encoding = "UTF-8"))
knitr::include_graphics(here::here("analysis", "figures", "007-raman-pca-with-photos.jpg"), error = FALSE)
```

@@

```{r}
#| label: fig-xrf-violin-plot
#| cache: true
#| fig-width: 7.87
#| fig-asp: 1
#| out-width: "70%"
#| fig-align: center
#| fig-cap: "Violin plots of the XRF values per element. Median and interquartile range is represented with a boxplot, with the half-violin showing kernel density estimation. Number of values above LOD is noted by n. Of particular note is the bimodal distribution of Si values. Y and Zn were left out as they had less than 30 values above LOD"

@@

```{r}
#| label: fig-xrf-pca
#| cache: true
#| fig-cap: "XRF PCA score plots: a) Score plot featuring the full dataset, cluster 1 includes all samples with Si > 90%, while the remaining samples seem to cluster based on Fe and Zr content. This results in most of the darker materials clustering towards the bottom of the plot. b) Score plot with > 90% Si samples excluded. Similar to plot A there is a separation based on Fe and Zr values, with darker material towards the bottom of the plot. Light-to-colourless samples remain largely together in cluster 3. Cluster 1 stands out by featuring slightly lower Si, Fe, Ti, and Zr values, but higher values in Al, Ca, K, and Mg"

knitr::include_graphics(here::here("analysis", "figures", "009-xrf-pca-with-photos.jpg"), error = FALSE)
```

@@

```{r}
#| label: fig-nir-xrf-kmeans
#| cache: true
#| fig-width: 10
#| fig-asp: 0.687
#| out-width: "70%"
#| fig-cap: "NIR PCA score plot for PC 1-2, with the K-means cluster from XRF PCA model B visualised in the colour of the points. Red points are samples that were excluded in the XRF PCA. The trend separating dark-light samples in the NIR PCA mostly keeps cluster 2-3 separated, showing some overlap in the results, but cluster 1 is fairly evenly distributed in the score plot. This means that whatever differences can be discerned in the compositional data does not seem to be detectable using the ASD"

>>>>

---
title: "Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
  - Johan Linderholm:
      email: johan.linderholm@umu.se
      institute: [UmU]
      correspondence: false
  - Paul Geladi:
      email: paul.geladi@slu.se
      institute: [SLU]
      correspondence: false
  - Philip I. Buckland:
      email: philip.buckland@umu.se
      institute: [UmU]
      correspondence: false
institute:
  - UmU: Environmental Archaeology Lab, Department of Historical, Philosophical and Religious studies, Umeå University, Umeå, Sweden
  - SLU: Department of Forest Biomaterials and Technology, Swedish University of Agricultural Sciences, Umeå, Sweden 
title-block-published: "Last updated"  
date: now
date-format: long
crossref: 
  fig-title: '**Fig.**'
  fig-labels: '**arabic**'
  fig-prefix: Fig.
format:
  pdf:
    documentclass: scrartcl
    fig-pos: "H"
    geometry: margin = 2.54cm 
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 300
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/quaternary-science-reviews.csl" # Insert path for the bib-style
abstract: |
  Northern Fennoscandia represents a geologically complex region affected by both glacial and postglacial processes. Provenance of locally available geological material thus becomes complex and fraught with uncertainty. Quartzite was a key material type utilized by hunter-gatherers in Northern Sweden around the 4 000 – 2 000 BP period, and is thus critical to the understanding of raw material procurement and material flow within the region. However, there is a severe lack of methodological development in the characterization of these materials. 125 points and preforms made from quartz/quartzite were sampled from 46 archaeological sites along the upper Ångerman river valley in Västerbotten, Sweden. The material has been analysed non-destructively using three separate portable spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence). Evaluation of the spectra and exploratory data analysis using Principal Component Analysis demonstrates detectable differences in the material that likely stem from diagenetic/paragenetic origin. The presence of graphite, muscovite and biotite could likewise provide information on the material’s metamorphic grade. In addition to reaffirming the potential of field-based screening instrumentation, these results will benefit future surveys of geological sources in the region. They also indicate potential for the construction of a predictive model that could classify the quartzite based on its chemical characteristic. Such a model would prove useful in future spatial analysis and testing of models of raw material management.
keywords: |
  Hunter-Gatherer; Lithics; Provenance; Quartzite; Vibrational spectroscopy; Scandinavia
---

@@

<!-- The following code chunk defines some general settings how code chunks should behave. -->
```{r setup, echo = FALSE}
library(officer)
library(officedown)
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300,
  encoding = "UTF-8"
)
```

@@

This study includes spectroscopic surface measurements collected on bifacial points (probable arrowheads and spear points) made of quartz and quartzite. The points have been recovered from archaeological sites during excavations and surveys, including stray finds, in Västerbotten County, Sweden (@fig-distribution-map & @tbl-point-data). The sites are distributed along the Ångerman river valley, from Lake Kultsjön in the north, towards the town of Åsele in the south. The dataset consists of `r length(tot_uniq_samples)` points, point fragments and preforms stored in the collections at Västerbottens museum (www.vbm.se) and have been analysed using NIRS, Raman scattering and XRF.

@@

\linebreak
<!--
```{r}
#| label: fig-distribution-map
#| cache: false
#| cache-lazy: false
#| fig-cap: "Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden. The glacial disturbance is visible in the widely distributed moraine. Tectonic symbology after [@gee_stephens20]. Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet"

knitr::include_graphics(here::here("analysis", "figures", "site_distribution_v1.jpg"), error = FALSE)
```
-->

![Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden. The glacial disturbance is visible in the widely distributed moraine. Tectonic symbology after Gee and Stephens (2020). Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet](../figures/site_distribution_v1.jpg){#fig-distribution-map width=100%, height=100%}

@@

\linebreak
<!--
```{r}
#| label: fig-artefacts
#| cache: false
#| cache-lazy: false
#| fig-cap: "Examples of the bifacial points studied. The material is heterogeneous in terms of colour, and the surface of the artefacts themselves is not uniform. Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation. Length varies between 12 – 138 mm, width between 8 – 67 mm, and thickness between 3 – 76 mm"

knitr::include_graphics(here::here("analysis", "figures", "figure_2.jpg"), error = FALSE)
```
-->

![Examples of the bifacial points studied. The material is heterogeneous in terms of colour, and the surface of the artefacts themselves is not uniform. Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation. Length varies between 12 – 138 mm, width between 8 – 67 mm, and thickness between 3 – 76 mm](../figures/figure_2.jpg){#fig-artefacts width=100%, height=100%}

@@

```{r}
#| label: tbl-point-data
#| cache: false
#| cache-lazy: false
#| tbl-cap: "Table of artefact location, material and type"

suppressPackageStartupMessages(library(flextable))

#custom function for fitting table to page width, see Sarah's answer on
#https://stackoverflow.com/questions/57175351/flextable-autofit-in-a-rmarkdown-to-word-doc-causes-table-to-go-outside-page-mar
FitFlextableToPage <- function(ft, pgwidth = 6){
  ft_out <- ft %>% autofit()
  ft_out <- width(ft_out, width = dim(ft_out)$widths*pgwidth /(flextable_dim(ft_out)$widths))
  return(ft_out)
}

metadata.csv <- 
  metadata.csv %>% 
  arrange(site_id)

#set table functions
set_flextable_defaults(
  theme_fun = theme_vanilla, 
  font.size = 10, 
  padding = 6,
  background.color = "#FFFFFF")

#columns to select
columns <- 
  c("sample_id", "site_id", "latitude", "longitude", "parish", "material", "find_context")
#names for table header
names <- 
  c("Sample ID", "Site", "Latitude", "Longitude", "Parish", "Material", "Find context")

#border options
inner_border = officer::fp_border(color="gray", width = 1)
header_border = officer::fp_border(color="black", width = 2)
body_border = officer::fp_border(color="black", width = 1)

ft <- 
  #select data
  flextable(metadata.csv, col_keys = columns) %>% 
  #set header names
  set_header_df(mapping = data.frame(keys = columns, values = names, stringsAsFactors = FALSE), key = "keys") %>% 
  #bold header font
  bold(bold = TRUE, part = "header") %>% 
  #align header names center
  align(align = "center", part = "all") %>% 
  #set borders and colors 
  border_inner_h(part = "all", border = inner_border ) %>% 
  border_inner_v(part = "all", border = inner_border ) %>% 
  border_outer(part = "body", border = body_border) %>% 
  border_outer(part = "footer", border = body_border) %>% 
  border_outer(part = "header", border = header_border) %>% 
  #use function to fit to page
  FitFlextableToPage

ft
```

@@

```{r}
#| label: fig-nir-peak-summary
#| cache: true
#| fig-width: 7.87
#| fig-asp: 1
#| out-width: "100%"
#| fig-align: center
#| fig-cap: "The main features in the NIR spectra can be related to overtones and vibration combinations of OH groups. The 1 400 nm and 1 900 nm peaks are indicative of molecular water, and common in quartz-based materials. The 2 200 nm and 2 350 nm bands indicate metal-OH combinations, where a more intense 2 200 nm band has typically been related to the presence of AlOH, and a more intense 2 350 nm band MgOH. Although difficult to detect in the non-processed spectra, weaker features also appear at 1 065 nm and 1 300 nm, that may be the result of ferrous iron (Fe^2+^)"

@@

\linebreak
<!--
#run this code chunk to display PCA plot without photos
```{r}
#| label: fig-nir-pca
#| cache: true
#| fig-width: 7.87
#| fig-asp: 1
#| out-width: "100%"
#| fig-align: center
#| fig-cap: "NIR PCA score plot for PC 1-3: a) dark and light material is separated out along PC 1, due to absorbance, with white and colourless material separating along PC 2, b) PC 3 adds little new information, mainly separating three of the light samples"

quietly(source(here::here("analysis", "paper","004-nir-pca.R"), encoding = "UTF-8"))

fig
```
-->

![NIR PCA score plot for PC 1-3: a) dark and light material is separated out along PC 1, due to absorbance, with white and colourless material separating along PC 2, b) PC 3 adds little new information, mainly separating three of the light samples](../figures/004-nir-pca-with-photos.jpg){#fig-nir-pca width=100%, height=100%}

@@

\linebreak
<!--
```{r}
#| label: fig-raman-spectra-summary
#| cache: true
#| fig-width: 7.87
#| fig-asp: 0.8
#| out-width: "100%"
#| fig-align: center
#| fig-cap: "The non-processed Raman spectra visualised by colour group. Note that the dark material in general features more intense peaks at wavenumber ~ 1 300 - 1 600 (cm^-1^)"

@@

```{r}
#| label: fig-raman-peak-ident
#| cache: true
#| fig-width: 7.87
#| fig-asp: 0.8
#| out-width: "100%"
#| fig-align: center
#| fig-cap: "TAveraged and baseline corrected Raman spectra for the light and dark groups. The features can be related to quartz and carbon structures in the material. The initial three peaks in 100 - 500 cm^-1^ range correspond to some of the more well-known vibration modes of quartz. The 1 287 cm^-1^ and 1 598 cm^-1^ peaks fall within what is referred to as the D and G band respectively, indicating the presence of carbon structures"

@@

\linebreak
<!--
#run this code chunk to display PCA plot without photos
```{r}
#| label: fig-raman-pca
#| cache: true
#| fig-width: 7.87
#| fig-asp: 1
#| out-width: "100%"
#| fig-align: center
#| fig-cap: "PCA score plot for Raman data. Objects that feature only quartz peaks are clustered towards the top left of the plot, whilst the darker material that feature dominant D/G peaks are distributed towards the right of the plot. In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster"

quietly(source(here::here("analysis", "paper", "007-raman-pca.R"), encoding = "UTF-8"))

fig
```
-->

![PCA score plot for Raman data. Objects that feature only quartz peaks are clustered towards the top left of the plot, whilst the darker material that feature dominant D/G peaks are distributed towards the right of the plot. In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster](../figures/007-raman-pca-with-photos.jpg){#fig-raman-pca width=100%, height=100%}

@@

```{r}
#| label: fig-xrf-violin-plot
#| cache: true
#| fig-width: 7.87
#| fig-asp: 1
#| out-width: "100%"
#| fig-align: center
#| fig-cap: "Violin plots of the XRF values per element. Median and interquartile range is represented with a boxplot, with the half-violin showing kernel density estimation. Number of values above LOD is noted by n. Of particular note is the bimodal distribution of Si values. Y and Zn were left out as they had less than 30 values above LOD"

@@

<!--
#run this code chunk to display PCA plot without photos
```{r}
#| label: fig-xrf-pca
#| cache: true
#| fig-width: 7.87
#| fig-asp: 1
#| out-width: "100%"
#| fig-align: center
#| fig-cap: "XRF PCA score plots: a) Score plot featuring the full dataset, cluster 1 includes all samples with Si > 90%, while the remaining samples seem to cluster based on Fe and Zr content. This results in most of the darker materials clustering towards the bottom of the plot. b) Score plot with > 90% Si samples excluded. Similar to plot A there is a separation based on Fe and Zr values, with darker material towards the bottom of the plot. Light-to-colourless samples remain largely together in cluster 3. Cluster 1 stands out by featuring slightly lower Si, Fe, Ti, and Zr values, but higher values in Al, Ca, K, and Mg"

fig2
```
-->

![XRF PCA score plots: a) Score plot featuring the full dataset, cluster 1 includes all samples with Si > 90%, while the remaining samples seem to cluster based on Fe and Zr content. This results in most of the darker materials clustering towards the bottom of the plot. b) Score plot with > 90% Si samples excluded. Similar to plot A there is a separation based on Fe and Zr values, with darker material towards the bottom of the plot. Light-to-colourless samples remain largely together in cluster 3. Cluster 1 stands out by featuring slightly lower Si, Fe, Ti, and Zr values, but higher values in Al, Ca, K, and Mg](../figures/009-xrf-pca-with-photos.jpg){#fig-xrf-pca width=100%, height=100%}

@@

```{r}
#| label: fig-nir-xrf-kmeans
#| cache: true
#| fig-width: 10
#| fig-asp: 0.687
#| out-width: "100%"
#| fig-cap: "NIR PCA score plot for PC 1-2, with the K-means cluster from XRF PCA model B visualised in the colour of the points. Red points are samples that were excluded in the XRF PCA. The trend separating dark-light samples in the NIR PCA mostly keeps cluster 2-3 separated, showing some overlap in the results, but cluster 1 is fairly evenly distributed in the score plot. This means that whatever differences can be discerned in the compositional data does not seem to be detectable using the ASD"
